Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-11 paper.Rmd
fix incorrect linetype in figure 1

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
 - Richard K. Morgan^[V-Dem, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: Cassy L. Dorff and Shahryar Minhas both provided helpful feedback on this project. All the code and several additional analyses can be found at our replication archive at https://github.com/rickmorgan2/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (2020).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  While their model is not very theoretical, it is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
#header-includes:
#  - \usepackage{setspace}\doublespacing
bibliography: 
  - "`r user <- Sys.info()[['user']]; if (user=='MooseJaw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
---

@@

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars remarkably accurately. Indeed they argue that this so-called theoretical model outperforms four other``more mechanical'' alternatives. Second, they claim that the integration of structure with process is better over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects. B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust, and a more precise test is undertaken. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

To examine the first question, B&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:
\begin{itemize}
\item Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
\item Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
\item Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government on one side and opposition or rebel actors on the other. These are directed, thus making for four total covariates.
\item CAMEO: counts for all CAMEO event codes totaling $1,159$ covariates, which are mostly zero for any country in any month. 
\item Average: unweighted average of the predictions from the four models briefly described above.
\end{itemize}

@@

This implies that the ROC curves were only smoothed in the referenced Figure 1. However, this is not the case. All AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation. ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like. Figure 1 shows our replication of both the estimated smoothed ROC curves from the B&S report (left-hand side) and the actual empirical ROC curves (right-hand side). The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.

![Replication of B&S Figure 1 with both smooth and non-smoothed ROC curves.](figures/figure1-replicated.png)
The specific predictions also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for an observed positive case is reached. For these ROC curves, and for that matter in the fundamental train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates. Notice also that the smoothing averages the left-most almost straight line with the right-most almost straight line in a monotonic way.

@@

1. B&S: "Of the approaches we test, the split-population analog is most promising \ldots "

@@

B&S advocate the use of theory to guide prediction. But ``theory'' as used by B&S is an ambiguous and undefined concept.  It is not a procedure.  They actually create a model with ten right-hand side variables that are supposed to capture a complicated repression-dissent dynamic. There is wide-ranging literature on this dynamic that could justify many specifications.  As such, their baseline comparison is an unfortunate standard bearer for strong theory.

@@

In contradistinction to the conclusions offered by B&S, we find that:
\begin{itemize}
\item The theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model.
\item The Average ensemble model and the CAMEO models outperform the escalation model in all instances.
\item Adding structural variables substantially improves the escalation model's performance.
\end{itemize}

>>>>

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com. corresponding author]
 - Richard K. Morgan^[V-Dem, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: Cassy L. Dorff and Shahryar Minhas both provided helpful feedback on this project. All the code and several additional analyses can be found at our replication archive at https://github.com/rickmorgan2/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (2020).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  While their model is not very theoretical, it is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
#header-includes:
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
---

@@

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars remarkably accurately. Indeed they argue that this so-called theoretical model outperforms four other "more mechanical" alternatives. Second, they claim that the integration of structure with process is better over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects. B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust, and a more precise test is undertaken. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

To examine the first question, B&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:

\begin{itemize}
\item Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
\item Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
\item Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government on one side and opposition or rebel actors on the other. These are directed, thus making for four total covariates.
\item CAMEO: counts for all CAMEO event codes totaling $1,159$ covariates, which are mostly zero for any country in any month. 
\item Average: unweighted average of the predictions from the four models briefly described above.
\end{itemize}

@@

This implies that the ROC curves were only smoothed in the referenced Figure 1. However, this is not the case. All AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation. ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like. Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B&S report (left-hand side) and the actual empirical ROC curves (right-hand side). The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.

\begin{figure}
\caption{Replication of B\&S Figure 1 with both smooth and non-smoothed ROC curves\label{fig:rocs}}
\centering
\includegraphics[width=.8\linewidth]{figures/figure1-replicated.png}
\end{figure}

The specific predictions also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for an observed positive case is reached. For these ROC curves, and for that matter in the fundamental train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates. Notice also that the smoothing averages the left-most almost straight line with the right-most almost straight line in a monotonic way.

@@

1. B&S: "Of the approaches we test, the split-population analog is most promising ..."

@@

B&S advocate the use of theory to guide prediction. But "theory" as used by B&S is an ambiguous and undefined concept.  It is not a procedure.  They actually create a model with ten right-hand side variables that are supposed to capture a complicated repression-dissent dynamic. There is wide-ranging literature on this dynamic that could justify many specifications.  As such, their baseline comparison is an unfortunate standard bearer for strong theory.

@@

In contradistinction to the conclusions offered by B&S, we find that:

\begin{itemize}
\item The theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model.
\item The Average ensemble model and the CAMEO models outperform the escalation model in all instances.
\item Adding structural variables substantially improves the escalation model's performance.
\end{itemize}
